Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-08-01 3.technique.tex
Fix some grammar issues

% !TeX spellcheck = en_US
\section{Race Detection by Mining Locking Rules}
\label{sec_technique}
To address the above challenges, we propose three key techniques. For {\em C1}, 
we propose an {\em alias-aware rule mining method} to automatically deduce 
locking rules. For {\em C2}, we propose a {\em lock-usage analysis} to filter 
out false data races by validating concurrency of code paths. For {\em C3}, we 
propose a pattern-based estimation to extract harmful races that can trigger 
memory or logical bugs such as null-pointer dereference and data inconsistency. 
We introduce them as follows:

\subsection{Alias-Aware Rule Mining Method}
\label{subsec_rule_mining}
The relationship between variables and locks are not well documented in OS 
kernels, but it can be inferred from the kernel code. Specifically, a variable 
is often protected by the lock stored in the same data structure. And thus if a 
variable is accessed after acquiring a lock existing in the same data structure 
in most cases, it is likely to be protected by the lock. Whether a variable and 
the protecting lock exist in the same data structure can be determined though 
an alias graph~\cite{Li:ASPLOS22, Kastrinis:CC18} by finding their common 
ancestor. Based on this insight, we propose an {\em alias-aware rule mining 
method} to deduce locking rules automatically. Moreover, with benefits from 
precise field-sensitive alias relationships of alias graph, our alias-aware 
rule mining method can find data structure filed and its protecting lock 
effectively.

\PP{Alias Graph.} It is an important data structure to infer relationships 
between variable and its protecting lock in our analysis, so we introduce it 
and its update first.

An alias graph is a 2-tuple $\mathit{G = \left<N, E\right>}$, where 
$\mathit{N}$ is a set of nodes, and each node $\mathit{n}$ represents an alias 
set that points to one abstract object. $\mathit{E}$ is a set of labeled edges. 
Each edge is labeled with a data structure field or a dereference operator 
``$\mathit{*}$'', which represents how an abstract object is accessed. In an 
alias node, a variable residing in a node followed by a sequence of edge
labels form an access path~\cite{Kastrinis:CC18, Cheng:PLDI00}. In this paper, 
we replace the variable in an access path with its structure name to represent 
a data structure field.

An alias graph is updated by handling four types of instructions that  
change alias relationships: MOVE($\mathit{v_1 = v_2}$), STORE ($\mathit{*v_2 = 
v_1}$), LOAD ($\mathit{v_1 = *v_2}$) and GEP ({$\mathit{v_1 = \&v_1->f}$}). We 
exploit {$\mathit{n_x}$} to represent the node whose representing alias set 
includes $\mathit{v_x}$, and introduce how the four types of instructions 
update alias graphs. For a MOVE operation, $\mathit{v_1}$ is moved from 
$\mathit{n_1}$ to $\mathit{n_2}$. After this operation, $\mathit{v_1}$ and  
$\mathit{v_2}$ are represented by the same node, which indicates they become 
aliases. For a STORE operation, the existing outgoing edge from $\mathit{n_2}$ 
is dropped first, and then a new edge labeled with $\mathit{*}$ from 
$\mathit{n_2}$ to $\mathit{n_1}$ is inserted. After this operation, 
$\mathit{v_1}$ and $\mathit{*v_2}$ are represented by the same node, which 
indicates they become aliases. For a LOAD operation, the analysis first finds 
the destination node of the edge that comes from $\mathit{n_2}$ and is labeled 
with $\mathit{*}$, and then move $\mathit{v_1}$ to the destination node. And 
after this operation, $\mathit{v_1}$ and $\mathit{*v_2}$ are represented by the 
same node, which indicates they become aliases. GEP operation is similar to 
LOAD, expect that the edge is labeled with a data structure field $\mathit{f}$, 
instead of a dereference operator $\mathit{*}$.

@@

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_alias_graph} shows a 
piece of driver-like source code and its alias graph. In this example, after an 
GEP (\&dev->lock) operation at Line 6, an edge labeled with {\tt lock} from 
node $\mathit{n_1}$ to node $\mathit{n_4}$ is inserted. Similarly, an edge 
labeled with {\tt flag} from node $\mathit{n_1}$ to node $\mathit{n_2}$ is 
inserted after Line 7. At last, an edge labeled with a dereference operator 
($\mathit{*}$) is inserted after the STORE (*fp = f) operation at Line 8. The 
final alias graph is shown in Figure~\ref{fig_demo_alias_graph}(b), and access 
paths are shown in the bottom left corner. Take node $\mathit{n_3}$ as an 
example, it can represent two fields, one is {\tt Dev.flag*}, and the other is 
{\tt int\_pointer*} (we exploit int\_pointer to represent a pointer points to 
an integer, and regard it as a data structure for convenience).

@@

Take the alias graph in Figure~\ref{fig_demo_alias_graph} as an example, {\tt 
\&dev->flag} is represented by node $\mathit{n_2}$, and {\tt \&dev->lock} is 
represented by node $\mathit{n_4}$. The two nodes have a common ancestor 
$\mathit{n_1}$, and thus the accessed variable {\tt \&dev->flag} and the 
protecting lock {\tt \&dev->lock} can be inferred to exist in the same data 
structure (namely {\tt Dev}). Therefore, the structure field Dev.flag is likely 
to be protected the lock stored in the structure field Dev.lock.

@@

Based on the alias graph, our alias-aware rule mining method performs an 
inter-procedural path-based~\cite{Li:ASPLOS22}, field-sensitive and alias-aware 
analysis to effectively mine locking rules about whether a specific variable 
should be protected by a lock and which lock is required. Overall, our 
alias-aware rule mining method has three main stages shown in 
Figure~\ref{fig_workflow}. In Stage 1, it extract key variable that may be 
protected by a lock. In Stage 2, it collect all accesses to key fields as well 
as code paths these accesses exist in. In Stage 3, it deduce locking rules by 
calculate the proportion of field accesses protected by a specific lock in all 
field accesses.

\PP{S1: Key-field extraction.} The OS Kernel has a large code base with 
numerous variables. However, only a small part of variables should be protected 
by a specific lock, and thus collecting all variable accesses can introduce 
much unnecessary overhead. Generally, only a few field accesses can miss 
necessary protecting lock, due to carelessness of developers. Based on this 
insight, our analysis first extracts key fields that may need to be protected 
by a specific lock, by performing a lock-set analysis to find whether a given 
variable is accessed after acquiring a lock in the same data structure as the 
accessed field in any code path.

@@

Figure~\ref{fig_pseudocode_lock_set} shows the lock-set analysis based on alias 
graph. The analysis first gets the operand of the instruction (Line 1), and 
then get the node of the operand from the alias graph (Line 2). If the 
instruction is a lock operation, the node of the operand is inserted into the 
lock set (Lines 3-4). Otherwise, if the instruction is an unlock operation, the 
node of the operand is removed from the lock set (Lines 5-6).

@@

Figure~\ref{fig_pseudocode_field_extract} shows the pseudocode to extract key 
fields that may be protected by a specific lock, based on lock-set analysis and 
the alias graph. The analysis starts from each function without a caller 
function (Lines 16-26) and performs a path-based analysis~\cite{Li:ASPLOS22} 
(Lines 17-25). For each instruction in the code path, the analysis first update 
the alias graph according to the instruction with the four operations (MOVE, 
STORE, LOAD and GEP) (Line 1), and then performs a lock-set analysis (Line 2) 
to get all acquired locks. After updating the alias graph and the lock set, if 
the instruction is a write or a read, the analysis first gets the node of the 
operand with the new alias graph (Lines 4-5). And then, for each node of 
lock variable in the lock set, the analysis uses the alias graph to extract the 
protected data structure field (Lines 6-12). If the field is not NULL, it is 
inserted into the set of key fields (Lines 9-11). For inter-procedural 
analysis, the analysis copies function definition into its call sites and this 
operation is not shown in Figure~\ref{fig_pseudocode_field_extract} for 
convenience.

\PP{S2: Field-access collection.} With the key fields extracted in Stage 1, the 
analysis can only collects accesses to the variables that is protected in some 
cases, because if a variable is never protected by any lock when is accessed, 
it is less likely to be shared by different threads, and thus can not introduce 
no concurrency issue.

@@

Figure~\ref{fig_pseudocode_access_collect} shows the pseudocode to collect all 
accesses to key fields. Similarly to the key-field extraction, this stage also 
performs a path-based analysis. For each instruction in the code path (Lines 
39-45), the analysis first updates the alias graph and the lock set according 
to the handled instruction (Lines 1-2). And then, the access type (either a 
write or a read) is set according to the instruction (Lines 5-8). However, if 
the instruction is not an access operation, the function returns with a NULL 
value for the field access (Line 10). Otherwise, the analysis first gets the 
node of the operand with the new alias graph (Lines 12-13), and then for each 
node of the lock variable in the lock set, the analysis uses the alias graph 
toextract the fields that the accessed variable and the lock variable stored in 
(Lines 16-17). If the fields are found, the field access is returned with the 
found fields (Lines 18-22). If the fields are not found, the accessed variable  
does not protected by any lock. In this case, if the accessed variable is 
stored in a key field, the field access is returned with a NULL lock variable 
(Lines 25-30).

\PP{S3: Locking-rule deduction.} After collecting all accesses to key fields, 
the analysis deduces locking rules based on statistics. However, on the one 
hand, distinguishing different accesses to the same data structure field by 
different program sites is not fine-grained enough, because the access to a 
variable and the acquirement to a specific lock are often packaged in a special 
function, and all accesses under different calling context are regarded as the 
same access. On the other hand, distinguishing different accesses by different 
code paths can also suffer from inaccuracy when a function contains many branch 
statement and cause numerous code paths for the same variable access. Based on 
these consideration, the analysis distinguishes different accesses to the same 
data structure field by different calling contexts. Specifically, given a key 
field {\em f} and a lock {\em l}, the analysis first finds all access to {\em 
f} with lock {\em l} from the collected field accesses in Stage 2, and gets the 
number {\em locked\_access\_num} by counting different calling contexts 
extracted from the code paths of these field accesses. Then, the analysis finds 
all access to {\em f} (no matter which lock is accessed), and gets the number 
{\em all\_access\_num} in the same way as {\em locked\_access\_num}. If the 
ratio {\em locked\_access\_num}/{\em all\_access\_num} is larger than a given 
threshold, and these is at least one write access to {\em f}, the data 
structure field {\em f} is inferred to be protected by the lock {\em l}. After 
deducting locking rules, the analysis detecting data races by checking whether 
a field access validates any locking rules.

@@

\noindent{\textbf{\em Example.}} We use an example in 
Figure~\ref{fig_demo_rule_mining} to illustrate how to mine locking rules 
with the three stages. The analysis first perform a path-based analysis to 
extract key fields. Take the code path {\tt Line 12, Line 6, Line 7, Line 8, 
Line 9} as an example. The analysis updates the alias graph by handling two GEP 
operations ({\em \&dev->lock} and {\em fp = dev->flag}) and a STORE operation 
({\em *fp = f}). The final alias graph is shown in 
Figure~\ref{fig_demo_rule_mining}(b). At the same time of alias analysis, the 
analysis records the acquired lock {\em Dev.lock} into the lock set after the 
lock instruction at Line 6. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus {\em Dev.flag} is a key field. Similarly, {\em 
Dev.flag*} is also a key field, due to the write instruction at Line 8.

@@

After collecting all field accesses, the analysis finds that the fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed under three calling contexts, and 
two of them are protected by the lock {\em Dev.lock}. And thus the two fields 
{\em Dev.flag} and {\em Dev.flag*} is deduce to be protected by the lock {\em 
Dev.lock} (assume the threshold of the ratio {\em locked\_access\_num} / 
{\em all\_access\_num} is 0.6). However, the accesses at Lines 16 and 17 are 
not protected by {\em Dev.lock} and thus causing two data races.

\subsection{Lock-Usage Analysis}
\label{subsec_lock_usage_analysis}
Our analysis takes functions that has no caller function as the entry as 
existing work~\cite{Li:ASPLOS22}, and performs a path-based analysis. It 
assumes that all codes reaching from entry functions can execute concurrently 
to reduce false negatives. However, this assumption is too conservative and can 
introduce many false positives because not all entry functions can execute 
concurrently in fact. We observe that each kernel module has an initialization 
phase, after which many function can be called concurrently by other modules. 
When performing initialization, the kernel module serially initialized the lock 
and prepare other data for subsequent operations. And thus functions with the 
lock initialization and functions called by them tend not to be executed 
concurrently. Based on the observation, we propose a lock-usage analysis to 
filter out false positives caused by code that can not execute concurrently.

@@

Our lock-usage analysis uses the union-find set~\cite{Galler:ACM64} to get all 
functions that reachable from the lock initialization function such as {\em 
mutex\_init()} and {\em spin\_lock\_init()}. 
Figure~\ref{fig_pseudocode_lock_usage} shows the pseudocodes of the lock-usage 
analysis. For each analyzed function in the OS kernel, the analysis first gets 
all the called function of it (Lines 3-4), and then union the analyzed function 
and the called function to update the union-find set (Line 5). After processing 
all the call instructions, the analysis judging whether a function executes in 
the initialization phase by checking whether it exist in the same set with the 
given lock initialization function (Lines 9-11).

@@

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_lock_usage} shows a false 
data race filtered out by our lock-usage analysis in the Linux DRM scheduler. 
For example in {\em drm\_sched\_entity\_set\_priority()}, the accesses to {\em 
entity->priority} is protected by the lock {\em entity->rq\_lock} in most 
cases. Therefore, {\em entity-priority} is deduced to be protected by the lock 
{\em entity->rq\_lock}. But in {\em drm\_sched\_entity\_init()}, {\em 
entity->priority} is written without acquiring the lock {\em entity->rq\_lock} 
and thus introducing a possible data race. However, the lock initialization 
function {\em spin\_lock\_init()} is called at Line 86 by {\em 
drm\_sched\_entity\_init()}. And thus {\em drm\_sched\_entity\_init()} is 
inferred to execute in the module initialization phase and can not execute 
concurrently with other function, and this is a false data race.

\subsection{Pattern-Based Estimation}
\label{subsec_estimation}
Many data races are benign or introduced by developers deliberately to improve 
the kernel performance. They can not cause memory or logic bugs, and thus 
developers are unwilling to put effort into repairing them. We observe that 
harmful data races match some typical patterns, and propose a {\em 
pattern-based estimation} to extract data races that can cause memory or logic 
bugs.

@@

\begin{itemize}
	\item \PP{Null-pointer dereference.} The two accesses to {\em dev->event} 
	are not protected by {\em dev->lock}. This can cause a null-pointer 
	dereference if {\em dev->event} is set to NULL right after the condition of 
	the if statement is checked to be true. To recognize such pattern, the 
	analysis first locates the data race, if it is checked by an if statement 
	and then dereferenced, a possible null-pointer dereference can occur.
	%\item \PP{Infinite loop.} The condition of the while statement does not 
	%protected by {\em dev->lock}. This can cause a infinite loop when {\em 
	%dev->cnt} is assigned with a non-zero value every time before it is 
	%checked 
	%by the while statement. The analysis detects such bugs by judge whether 
	%the 
	%data race acts as a condition of a loop statement.
	\item \PP{Data inconsistency.} Two different fields of the same data 
	structure are accessed without acquiring {\em dev->lock}. This can cause a 
	data inconsistency when the data structure field {\em dev->clock} is 
	changed by another thread right after the access to {\em dev->clock->min}. 
	To recognize such pattern, the analysis detect whether multiple fields of 
	the same data structure are accessed without acquiring the protecting lock.
	\item \PP{Unprotected write.} The write access to {\em dev->event} is not 
	protected by {\em dev->lock}. This is dangerous because the value {\em 
	dev->event} can be modified at any time when other threads access it, and 
	thus introduce unpredictable behavior. The analysis detect such pattern by 
	judge whether a data race occurs in a write access.
\end{itemize}

>>>>

% !TeX spellcheck = en_US
\section{Race Detection by Mining Locking Rules}
\label{sec_technique}
To address the above challenges, we propose three key techniques. For {\em C1}, 
we propose an {\em alias-aware rule mining method} to automatically deduce 
locking rules. For {\em C2}, we propose a {\em lock-usage analysis} to filter 
out false data races by validating the concurrency of kernel code. For {\em 
C3}, we propose a pattern-based estimation to extract harmful races that can 
trigger memory or logical bugs such as null-pointer dereference and data 
inconsistency. We introduce them as follows:

\subsection{Alias-Aware Rule Mining Method}
\label{subsec_rule_mining}
The relationship between variables and locks is not well documented in OS 
kernels, but it can be inferred from the kernel code. Specifically, a variable 
is often protected by the lock stored in the same data structure. And thus if a 
variable is accessed after acquiring a lock existing in the same data structure 
in most cases, it is likely to be protected by the lock. Whether a variable and 
the protecting lock exist in the same data structure can be determined through 
an alias graph~\cite{Li:ASPLOS22, Kastrinis:CC18} by finding their common 
ancestor. Based on this insight, we propose an {\em alias-aware rule mining 
method} to deduce locking rules automatically. Moreover, with benefits from 
precise field-sensitive alias relationships of alias graph, our alias-aware 
rule mining method can find data structure filed and its protecting lock 
effectively.

\PP{Alias Graph.} It is an important data structure to infer relationship 
between variable and its protecting lock in our analysis, so we introduce it 
and its update first.

An alias graph is a 2-tuple $\mathit{G = \left<N, E\right>}$, where 
$\mathit{N}$ is a set of nodes, and each node $\mathit{n}$ represents an alias 
set that points to one abstract object. $\mathit{E}$ is a set of labeled edges. 
Each edge is labeled with a data structure field or a dereference operator 
``$\mathit{*}$'', which represents how an abstract object is accessed. In an 
alias node, a variable residing in a node followed by a sequence of edge
labels forms an access path~\cite{Kastrinis:CC18, Cheng:PLDI00}. In this paper, 
we replace the variable in an access path with its structure name to represent 
a data structure field.

An alias graph is updated by handling four types of instructions that  
change alias relationships: MOVE($\mathit{v_1 = v_2}$), STORE ($\mathit{*v_2 = 
v_1}$), LOAD ($\mathit{v_1 = *v_2}$) and GEP ({$\mathit{v_1 = \&v_1->f}$}). We 
exploit {$\mathit{n_x}$} to represent the node whose representing alias set 
includes $\mathit{v_x}$, and introduce how the four types of instructions 
update alias graphs. For a MOVE operation, $\mathit{v_1}$ is moved from 
$\mathit{n_1}$ to $\mathit{n_2}$. After this operation, $\mathit{v_1}$ and  
$\mathit{v_2}$ are represented by the same node, which indicates they become 
aliases. For a STORE operation, the existing outgoing edge from $\mathit{n_2}$ 
is dropped first, and then a new edge labeled with $\mathit{*}$ from 
$\mathit{n_2}$ to $\mathit{n_1}$ is inserted. After this operation, 
$\mathit{v_1}$ and $\mathit{*v_2}$ are represented by the same node, which 
indicates they become aliases. For a LOAD operation, the analysis first finds 
the destination node of the edge that comes from $\mathit{n_2}$ and is labeled 
with $\mathit{*}$, and then moves $\mathit{v_1}$ to the destination node. And 
after this operation, $\mathit{v_1}$ and $\mathit{*v_2}$ are represented by the 
same node, which indicates they become aliases. GEP operation is similar to 
LOAD, expect that the edge is labeled with a data structure field $\mathit{f}$, 
instead of a dereference operator $\mathit{*}$.

@@

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_alias_graph} shows a 
piece of driver-like source code and its alias graph. In this example, after a 
GEP (\&dev->lock) operation at Line 6, an edge labeled with {\tt lock} from 
node $\mathit{n_1}$ to node $\mathit{n_4}$ is inserted. Similarly, an edge 
labeled with {\tt flag} from node $\mathit{n_1}$ to node $\mathit{n_2}$ is 
inserted after Line 7. At last, an edge labeled with a dereference operator 
($\mathit{*}$) is inserted after the STORE (*fp = f) operation at Line 8. The 
final alias graph is shown in Figure~\ref{fig_demo_alias_graph}(b), and access 
paths are shown in the bottom left corner. Take node $\mathit{n_3}$ as an 
example, it can represent two fields, one is {\tt Dev.flag*}, and the other is 
{\tt int\_pointer*} (we exploit int\_pointer to represent a pointer points to 
an integer, and regard it as a data structure for convenience).

@@

Take the alias graph in Figure~\ref{fig_demo_alias_graph} as an example, {\tt 
\&dev->flag} is represented by node $\mathit{n_2}$, and {\tt \&dev->lock} is 
represented by node $\mathit{n_4}$. The two nodes have a common ancestor 
$\mathit{n_1}$, and thus the accessed variable {\tt \&dev->flag} and the 
protecting lock {\tt \&dev->lock} can be inferred to exist in the same data 
structure (namely {\tt Dev}). Therefore, the structure field Dev.flag is likely 
to be protected by the lock stored in the structure field Dev.lock.

@@

Based on the alias graph, our alias-aware rule mining method performs an 
inter-procedural path-based~\cite{Li:ASPLOS22}, field-sensitive and alias-aware 
analysis to effectively mine locking rules about whether a specific variable 
should be protected by a lock and which lock is required. Overall, our 
alias-aware rule mining method has three main stages shown in 
Figure~\ref{fig_workflow}. In Stage 1, it extracts key variables that may be 
protected by a lock. In Stage 2, it collects all accesses to key fields as well 
as code paths these accesses exist in. In Stage 3, it deduces locking rules by 
calculating the proportion of field accesses protected by a specific lock in 
all field accesses.

\PP{S1: Key-field extraction.} The OS Kernel has a large code base with 
numerous variables. However, only a small part of variables should be protected 
by a specific lock, and thus collecting all variable accesses can introduce 
much unnecessary overhead. Generally, only a few field accesses can miss 
necessary protecting locks, due to the carelessness of developers. Based on 
this insight, our analysis first extracts key fields that may need to be 
protected by a specific lock, by performing a lock-set analysis to find whether 
a given variable is accessed after acquiring a lock in the same data structure 
as the accessed field in any code path.

@@

Figure~\ref{fig_pseudocode_lock_set} shows the lock-set analysis based on alias 
graph. The analysis first gets the operand of the instruction (Line 1), and 
then gets the node of the operand from the alias graph (Line 2). If the 
instruction is a lock operation, the node of the operand is inserted into the 
lock set (Lines 3-4). Otherwise, if the instruction is an unlock operation, the 
node of the operand is removed from the lock set (Lines 5-6).

@@

Figure~\ref{fig_pseudocode_field_extract} shows the pseudocode to extract key 
fields that may be protected by a specific lock, based on lock-set analysis and 
the alias graph. The analysis starts from each function without a caller 
function (Lines 16-26) and performs a path-based analysis~\cite{Li:ASPLOS22} 
(Lines 17-25). For each instruction in the code path, the analysis first update 
the alias graph according to the instruction with the four operations (MOVE, 
STORE, LOAD and GEP) (Line 1), and then performs a lock-set analysis (Line 2) 
to get all acquired locks. After updating the alias graph and the lock set, if 
the instruction is a write or a read, the analysis first gets the node of the 
operand with the new alias graph (Lines 4-5). And then, for each node of the 
lock variable in the lock set, the analysis uses the alias graph to extract the 
protected data structure field (Lines 6-12). If the field is not NULL, it is 
inserted into the set of key fields (Lines 9-11). For inter-procedural 
analysis, the analysis copies function definition into its call sites, but this 
operation is not shown in Figure~\ref{fig_pseudocode_field_extract} for 
convenience.

\PP{S2: Field-access collection.} With the key fields extracted in Stage 1, the 
analysis only needs to collect accesses to the variables that are protected in 
some cases, because if a variable is never protected by any lock when is 
accessed, it is less likely to be shared by different threads, and thus can not 
introduce any concurrency issue.

@@

Figure~\ref{fig_pseudocode_access_collect} shows the pseudocode to collect all 
accesses to key fields. Similarly to the key-field extraction, this stage also 
performs a path-based analysis. For each instruction in the code path (Lines 
39-45), the analysis first updates the alias graph and the lock set according 
to the handled instruction (Lines 1-2). And then, the access type (either a 
write or a read) is set according to the instruction (Lines 5-8). However, if 
the instruction is not an access operation, the function returns a NULL value 
for the field access (Line 10). Otherwise, the analysis first gets the node of 
the operand with the new alias graph (Lines 12-13), and then for each node of 
the lock variable in the lock set, the analysis uses the alias graph to extract 
the fields that the accessed variable and the lock variable stored in (Lines 
16-17). If the fields are found, the field access is returned with the found 
fields (Lines 18-22). If the fields are not found, the accessed variable  are 
not protected by any lock. In this case, if the accessed variable is stored in 
a key field, the field access is returned with a NULL lock variable (Lines 
25-30).

\PP{S3: Locking-rule deduction.} After collecting all accesses to key fields, 
the analysis deduces locking rules based on statistics. However, on the one 
hand, distinguishing different accesses to the same data structure field by 
different program sites is not fine-grained enough, because the access to a 
variable and the acquirement to a specific lock are often packaged in a special 
function, and all accesses under different calling context are regarded as the 
same access. On the other hand, distinguishing different accesses by different 
code paths can also suffer from inaccuracy when a function contains many branch 
statement and cause numerous code paths for the same variable access. Based on 
this consideration, the analysis distinguishes different accesses to the same 
data structure field by different calling contexts. Specifically, given a key 
field {\em f} and a lock {\em l}, the analysis first finds all access to {\em 
f} with lock {\em l} from the collected field accesses in Stage 2, and gets the 
number {\em locked\_access\_num} by counting different calling contexts 
extracted from the code paths of these field accesses. Then, the analysis finds 
all access to {\em f} (no matter which lock is accessed), and gets the number 
{\em all\_access\_num} in the same way as {\em locked\_access\_num}. If the 
ratio {\em locked\_access\_num}/{\em all\_access\_num} is larger than a given 
threshold, and there is at least one write access to {\em f}, the data 
structure field {\em f} is inferred to be protected by the lock {\em l}. After 
deducting locking rules, the analysis detects data races by checking whether a 
field access validates any locking rules.

@@

\noindent{\textbf{\em Example.}} We use an example in 
Figure~\ref{fig_demo_rule_mining} to illustrate how to mine locking rules 
with the three stages. The analysis first performs a path-based analysis to 
extract key fields. Take the code path {\tt Line 12, Line 6, Line 7, Line 8, 
Line 9} as an example. The analysis updates the alias graph by handling two GEP 
operations ({\em \&dev->lock} and {\em fp = dev->flag}) and a STORE operation 
({\em *fp = f}). The final alias graph is shown in 
Figure~\ref{fig_demo_rule_mining}(b). At the same time as alias analysis, the 
analysis records the acquired lock {\em Dev.lock} into the lock set after the 
lock instruction at Line 6. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus {\em Dev.flag} is a key field. Similarly, {\em 
Dev.flag*} is also a key field, due to the write instruction at Line 8.

@@

After collecting all field accesses, the analysis finds that the fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed under three calling contexts, and 
two of them are protected by the lock {\em Dev.lock}. And thus the two fields 
{\em Dev.flag} and {\em Dev.flag*} are deduced to be protected by the lock {\em 
Dev.lock} (assume the threshold of the ratio {\em locked\_access\_num} / 
{\em all\_access\_num} is 0.6). However, the accesses at Lines 16 and 17 are 
not protected by {\em Dev.lock} and thus causing two data races.

\subsection{Lock-Usage Analysis}
\label{subsec_lock_usage_analysis}
Our analysis takes functions that have no caller function as the entry as 
existing work~\cite{Li:ASPLOS22}, and performs a path-based analysis. It 
assumes that all codes reaching from entry functions can execute concurrently 
to reduce false negatives. However, this assumption is too conservative and can 
introduce many false positives because not all entry functions can execute 
concurrently in fact. We observe that each kernel module has an initialization 
phase, after which many functions can be called concurrently by other modules. 
When performing initialization, the kernel module serially initialized the lock 
and prepare other data for subsequent operations. And thus functions with the 
lock initialization and functions called by them tend not to be executed 
concurrently. Based on the observation, we propose a lock-usage analysis to 
filter out false positives caused by code that can not execute concurrently.

@@

Our lock-usage analysis uses the union-find set~\cite{Galler:ACM64} to get all 
functions that are reachable from the lock initialization function such as {\em 
mutex\_init()} and {\em spin\_lock\_init()}. 
Figure~\ref{fig_pseudocode_lock_usage} shows the pseudocodes of the lock-usage 
analysis. For each analyzed function in the OS kernel, the analysis first gets 
all the called functions of it (Lines 3-4), and then union the analyzed  
function and the called function to update the union-find set (Line 5). After 
processing all the call instructions, the analysis judges whether a function 
executes in the initialization phase by checking whether it exists in the same 
set with the given lock initialization function (Lines 9-11).

@@

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_lock_usage} shows a false 
data race filtered out by our lock-usage analysis in the Linux DRM scheduler. 
For example in {\em drm\_sched\_entity\_set\_priority()}, the accesses to {\em 
entity->priority} is protected by the lock {\em entity->rq\_lock} in most 
cases. Therefore, {\em entity-priority} is deduced to be protected by the lock 
{\em entity->rq\_lock}. But in {\em drm\_sched\_entity\_init()}, {\em 
entity->priority} is written without acquiring the lock {\em entity->rq\_lock} 
and thus introducing a possible data race. However, the lock initialization 
function {\em spin\_lock\_init()} is called at Line 86 by {\em 
drm\_sched\_entity\_init()}. And thus {\em drm\_sched\_entity\_init()} is 
inferred to execute in the module initialization phase and can not execute 
concurrently with other functions, and this is a false data race.

\subsection{Pattern-Based Estimation}
\label{subsec_estimation}
Many data races are benign or introduced by developers deliberately to improve 
kernel performance. They can not cause memory or logic bugs, and thus 
developers are unwilling to put effort into repairing them. We observe that 
harmful data races match some typical patterns, and propose a {\em 
pattern-based estimation} to extract data races that can cause memory or logic 
bugs.

@@

\begin{itemize}
	\item \PP{Null-pointer dereference.} The two accesses to {\em dev->event} 
	are not protected by {\em dev->lock}. This can cause a null-pointer 
	dereference if {\em dev->event} is set to NULL right after the condition of 
	the if statement is checked to be true. To recognize such a pattern, the 
	analysis first locates the data race, if it is checked by an if statement 
	and then dereferenced, a possible null-pointer dereference can occur.
	%\item \PP{Infinite loop.} The condition of the while statement does not 
	%protected by {\em dev->lock}. This can cause a infinite loop when {\em 
	%dev->cnt} is assigned with a non-zero value every time before it is 
	%checked 
	%by the while statement. The analysis detects such bugs by judge whether 
	%the 
	%data race acts as a condition of a loop statement.
	\item \PP{Data inconsistency.} Two different fields of the same data 
	structure are accessed without acquiring {\em dev->lock}. This can cause a 
	data inconsistency when the data structure field {\em dev->clock} is 
	changed by another thread right after the access to {\em dev->clock->min}. 
	To recognize such a pattern, the analysis detects whether multiple fields 
	of 	the same data structure are accessed without acquiring the protecting 
	lock.
	\item \PP{Unprotected write.} The write access to {\em dev->event} is not 
	protected by {\em dev->lock}. This is dangerous because the value {\em 
	dev->event} can be modified at any time when other threads access it, and 
	thus introduce unpredictable behavior. The analysis detects such a pattern 
	by judging whether a data race occurs in a write access.
\end{itemize}

### 2023-08-01 4.framework.tex
Fix some grammar issues

Based on the three key techniques in Section~\ref{sec_technique}, we develop a 
practical static approach named \sys, to detect data races in OS kernels, and 
estimate harmfulness of these data races. We have implemented \sys with 
Clang~\cite{clang} and Z3~\cite{z3}. \sys automatically mines locking rules, 
detects data races and estimates the harmfulness of detected data races. 
Figure~\ref{fig_architecture} shows the architecture of \sys, which has four 
phases:

@@

\PP{P3: Data race detection and false-positive filtering.} The data race 
detector detecting data races by checking whether a given access violates the 
rules mined by our locking-rule miner. After finding a possible data race, the 
data race filter uses our locking-usage analysis to determine whether it is a 
false positive. Besides, for a given data race, there may be multiple code 
paths from the entry function to its problematic instruction, and thus many 
repeated data races can be reported. To drop repeated data races, for a new 
possible data race, the data race filter checks whether its problematic 
instruction is identical to any existing data race. If so, this possible data 
race is regarded as repeated and dropped.

>>>>

Based on the three key techniques in Section~\ref{sec_technique}, we develop a 
practical static approach named \sys, to detect data races in OS kernels, and 
estimate the harmfulness of these data races. We have implemented \sys with 
Clang~\cite{clang} and Z3~\cite{z3}. \sys automatically mines locking rules, 
detects data races and estimates the harmfulness of detected data races. 
Figure~\ref{fig_architecture} shows the architecture of \sys, which has four 
phases:

@@

\PP{P3: Data race detection and false-positive filtering.} The data race 
detector detects data races by checking whether a given access violates the 
rules mined by our locking-rule miner. After finding a possible data race, the 
data race filter uses our locking-usage analysis to determine whether it is a 
false positive. Besides, for a given data race, there may be multiple code 
paths from the entry function to its problematic instruction, and thus many 
repeated data races can be reported. To drop repeated data races, for a new 
possible data race, the data race filter checks whether its problematic 
instruction is identical to any existing data race. If so, this possible data 
race is regarded as repeated and dropped.
